Accelerating parameter inference with graphics processing units

### 2019-02-03 paper/paper.tex
paper.tex: minor edits

Unless otherwise noted, all mass quantities described in this work are redshifted masses $m_{k,z}=(1+z)m_k$.

@@

 We perform a multi-stage iterative RIFT analysis of these signal with the
time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
spin.   For both sources, we adopt the fiducial distance prior $p(D_L) \propto D_L^2$,   selecting  a maximum distance
roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron star.
For the binary black hole we generate the signal with $f_{\rm min}=8 \unit{Hz}$ and analyze the signal with  $f_{\rm
  min}=10 \unit{Hz}$; for the binary neutron star, we generate the signal with $f_{\rm min}=20\unit{Hz}$ and anayze it
with $f_{\rm min}=23\unit{Hz}$.

@@

Binary neutron star models with more parameters like spin and tides require correspondingly larger numbers of points in the
  initial grid and per iteration.  Fortunately, the number of observationally significant and accessible dimensions is often
  substantially less than the a prior necessary dimensionality.  In practice,we use roughly $20\times$ more points per
  iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).    As a
  result, we can still achieve relatively rapid turnaround on a high-dimensional binary neutron star analysis.

@@

Due to our self-imposed resource constraints on $N_{\rm eval}=20$ and the number of GPUs, the first iteration is
resource-limited and requires
of order $(5000/20/100)\simeq
2.5$ to $(5000/20/48)\simeq
 5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
 evaluation in the first requires roughly 60 seconds (SpinTaylorT4) or  (SEOBNRv4ROM), on average.   Subsequent iterations proceed
 in  \editremark{timing}

@@

The overall code performance and thus latency can be further decreased substantially, notably by converting the Monte
Carlo random number generation and inner products to GPU-based operations.  In such a configuration, the marginalized
likelihood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm
  marg}(\lambda)$ for both binary black holes (which are Monte Carlo limited) and binary neutron stars (which are
inner-product limited).   We anticipate a further factor of roughly 10 reduction in overall evaluation time can be
achieved soon.  At that level of performance, a single 8-GPU machine with contemporary hardware could perform parameter inference less than 10
minutes.    Since binary compact objects intersect our past light cone only once every roughly 15 minutes, accounting
for all past history, such a configuration would be able to address low-latency parameter inference for the duration of
2nd-generation ground-based observing.
% based on Neval ~ 10^4

By allowing  models with higher-order modes to be used in low latency, our code can better-constrain the properties of low-mass
binaries with suitable amplitudes and orientations \editremark{citations to old work}, as demonstrated in Section
\ref{sec:sub:BNSFull}.  These tighter constraints could
better  inform low-latency source classification and hence multimessenger followup observations of compact binary mergers.

@@

\begin{acknowledgements}
RO'S is supported by NSF PHY-1707965 and PHY-1607520.
DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 
and New York State Urban Development Corporation, d/b/a Empire State Development, under contract No.\,AA289.
We thank the Computational Science Initiative at BNL for hosting GPU Hackathon 2018, during which a part of this work was conducted. We also thank Jeff Layton at NVIDIA for bringing CuPy to our attention.
\end{acknowledgements}

>>>>

@@

 We perform a multi-stage iterative RIFT analysis of these signal with the
time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
spin.   For both sources, we adopt the fiducial distance prior $p(D_L) \propto D_L^2$,   selecting  a maximum distance
roughly five times larger than the known source distance: 1Gpc for the binary black hole and 500 Mpc for the neutron
star.  We adopt a uniform prior on the component (redshifted) masses $m_{i,z}$ (and, when appropriate, spins
$\chi_{i,z}$).
Unless otherwise noted, all mass quantities described in this work are redshifted masses $m_{k,z}=(1+z)m_k$.
For the binary black hole we generate the signal with $f_{\rm min}=8 \unit{Hz}$ and analyze the signal with  $f_{\rm
  min}=10 \unit{Hz}$; for the binary neutron star, we generate the signal with $f_{\rm min}=20\unit{Hz}$ and anayze it
with $f_{\rm min}=23\unit{Hz}$.

@@

Binary neutron star models with more parameters like spin and tides require correspondingly larger numbers of points in the
  initial grid and per iteration.  Fortunately, the number of observationally significant and accessible dimensions is often
  substantially less than the a prior necessary dimensionality.  In practice,we use roughly $20\times$ more points per
  iteration for precessing massive BBH systems ($d=8$) or for spinning binary neutron stars with tides ($d=6$).    As a
  result, we can still achieve relatively rapid turnaround on a high-dimensional binary neutron star analysis even in
  resource-constrained environments.

@@

Due to our self-imposed resource constraints on $N_{\rm eval}=20$ and the number of GPUs, the first iteration is
resource-limited and requires
of order $(5000/20/100)\simeq
2.5$ to $(5000/20/48)\simeq
 5$ times as long as the first iteration of the zero-spin BNS analysis described above.  Each marginalized likelihood
 evaluation in the first requires roughly 60 seconds for both SpinTaylorT4 and SEOBNRv4ROM, on average.   Subsequent
 iterations are not resource-limited and complete in roughly 20 minutes, depending on hardware \editremark{timing}

@@

The overall code performance and thus latency can be further decreased substantially, notably by converting the Monte
Carlo random number generation and inner products to GPU-based operations.  In such a configuration, the marginalized
likelihood would carry out almost all calculations (except waveform generation) on a GPU, with minimal communication off
the board.  This configuration should further reduce the average time needed to compute $\ln {\cal L}_{\rm
  marg}(\lambda)$ for both binary black holes (which are Monte Carlo limited) and binary neutron stars (which are
inner-product limited).   We anticipate a further factor of roughly 10 reduction in overall evaluation time can be
achieved soon.  At that level of performance, a single 8-GPU machine with contemporary hardware could perform parameter inference less than 10
minutes.    Since binary compact objects intersect our past light cone only once every roughly 15 minutes, accounting
for all past history, such a configuration would be able to address low-latency parameter inference for the duration of
2nd-generation ground-based observing.
% based on Neval ~ 10^4
Alternatively, if larger resource pools are available in low latency, both the original and now GPU-accelerated RIFT  can
perform extremely rapid parameter inference if large iterations are performed completely in parallel (i.e., $N_{\rm
  eval} \simeq 1$).

By allowing  models with higher-order modes to be used in low latency, our code can better-constrain the properties of low-mass
binaries with suitable amplitudes and orientations, when circumstances allow.  These tighter constraints could
better  inform low-latency source classification and hence multimessenger followup observations of compact binary mergers.

@@

\begin{acknowledgements}
DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
JL is supported by NSF PHY-1707965.  
RO'S is supported by NSF PHY-1707965 and PHY-1607520.
Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 
and New York State Urban Development Corporation, d/b/a Empire State Development, under contract No.\,AA289.
We thank the Computational Science Initiative at BNL for hosting GPU Hackathon 2018, during which a part of this work was conducted. We also thank Jeff Layton at NVIDIA for bringing CuPy to our attention.
\end{acknowledgements}
